---
title: "Untitled"
output: pdf_document
---
##Processing the species abundances from Bracken for the TwinsUK 250. First read in the combined table and separate out the counts. Then divide by the geometric mean? Only?
```{r}
library(dplyr)
library(tidyr)
library(data.table)
library(tidyverse)
library(sctransform)
library(ggplot2)
library(COUNT)
```



```{r}
species <- read.delim2("/workdir/users/fnn3/twins_uk/taxa/bracken/combined_species_bracken.txt", header=T, sep="\t")
sab <- cbind(species$name, select(species, contains("num"))) #7,160 species
  sab <-   rename(sab, name = `species$name`)

sab2 <- sab[,-c(1)]

sab3 <- sab[rowMeans(sab[,-c(1)]!=0)>0.10,] #5,901 species left

```

##First step, divide all counts by the geometric mean to normalize for library size and compositionality
```{r}
sab$name <- gsub('\\s+', "_", sab$name)
tsab <- t(sab)

tsab <- setNames(data.frame(t(sab[,-1])), sab[,1])
tsab.m <- as.matrix(tsab)

#Get the geometric mean and add 1 to every value
gmean <-as.data.frame( sctransform:::row_gmean(tsab.m, eps=1))
names(gmean) <-c("gmean")

tsab2 <- merge(tsab, gmean, by=0 )
tsab2$Row.names <- gsub(".bracken_num","", tsab2$Row.names)
rownames(tsab2) <- tsab2[,1]
tsab2[,1] <- NULL

#Now to divide each value in each row by its own geometric mean
tsab.norm <- tsab2/tsab2$gmean

tsab.norm <- data.frame(BGI_ID=rownames(tsab.norm), tsab.norm)
tsab.norm$gmean <- NULL
write.table(tsab.norm, "/workdir/users/fnn3/twins_uk/taxa/abundance/twinsUK_species_abundance_gmeanNormalized.txt", sep="\t", quote = F)

#Remove rare data, 10%

tsab.norm2 <- tsab.norm[sapply(tsab.norm, function(x) mean(x==0) <= 0.9)] #5912 left
write.table(tsab.norm2, "/workdir/users/fnn3/twins_uk/taxa/abundance/twinsUK_species_abundance_gmeanNormalized_filt10.txt", sep="\t", quote = F, row.names=F)


#Now prepare them for CCA - remove the ten missing twins
withsnp <- read.table("/workdir/users/fnn3/twins_uk/BGI_SNP/BGI_twins_with_SNPs.txt", sep="\t", header=T)
tsab.norm240 <- tsab.norm2[tsab.norm2$BGI_ID %in% withsnp$bgi_id,]

write.table(tsab.norm240, "/workdir/users/fnn3/twins_uk/taxa/abundance/twinsUK240_species_abundance_gmeanNormalized_filt10.txt", sep="\t", quote = F, row.names=F)

```
###Plotting the species abundances to check for Tweedie. Normalized data
```{r}
msab <- apply(tsab.norm[,-1], 2, function(col)
  c("mean" = mean(col) , "pofzeros" = sum(col==0)/length(col))
  )   #mean abundance and proportion of zeros for each species aka 'col'

mean.ab <- msab[1,]
pofzeros.ab <- msab[2,]
std.abund2 <- apply(tsab.norm[,-c(1)], 2, function(vecc) sqrt(sum(vecc^2)))
log.var2 <- log(std.abund2^2)


plot(mean.ab, pofzeros.ab, main="Mean species abundance vs proportion of zeros") #looks weird...
plot(exp(-mean.ab^0.5/1.6), pofzeros.ab, main = "Tweedie expected proportion of zeros vs observed proportion of zero", ylim=c(0,1)) #looks ok?

pofnonzero <- 1 - pofzeros.ab
log.mean.abund2 <- log(mean.ab)

our.glm <- glm(pofnonzero ~ log.mean.abund2, family = binomial(link=cloglog), weights = rep( 250, 7160))

int <- our.glm$coefficients[1]
slope <- our.glm$coefficients[2]
optimal.transformation.of.mean <- exp(-mean.ab^slope/exp(-int)) #theoretically optimal: see email

pdf("/workdir/users/fnn3/twins_uk/taxa/plots/tweedie_transformation_mean_plot.pdf", height=5, width=10)
ggplot(data.frame('x'=optimal.transformation.of.mean, 'y'=pofzeros.ab , 'col'=log.var2), aes(x=x, y=y, colour=col))+
  geom_point(size=0.5, alpha=.5) + 
  xlab("Expected proportion of zeros") + ylab("Observed proportion of zeros") +
  scale_color_continuous(name="log(variance)")
dev.off()


#qplot(optimal.transformation.of.mean, pofzeros.abund2)
ggplot(data.frame('x'=optimal.transformation.of.mean, 'y'=pofzeros.ab, 'col'=log.var2), aes(x=x, y=y, colour=col))+
  geom_point(size=.5, alpha=.5)

qplot(log.mean.abund2, log.var2)+geom_abline(slope = 1, intercept = 0) + xlab("log mean") + ylab("log variance")
summary(lm(log.var2~log.mean.abund2))$coefficients[2,1]

pdf("/workdir/users/fnn3/twins_uk/taxa/plots/twins_uk_logmean_logvar.pdf", height=6, width=10)
ggplot(data.frame('x'=log.mean.abund2, 'y'=log.var2, 'col'=pofzeros.ab), aes(x=x, y=y, colour=col))+
  geom_point( alpha=.5) +
  ylab("log(var)") + xlab("log(mean)") +
  scale_color_continuous(name="Proportion of zeros") +
  geom_abline(intercept=0, slope=1)
dev.off()

plot(log.mean.abund2, log.var2, pch=20, xlim=c(0,8.5), ylim=c(0,34))#, pch=20, cex=.01)
abline(a=6, b=3.2, col='blue') # from zeros
plot(log.mean.abund2, log.var2, pch=20)#, pch=20, cex=.01)
abline(a=6, b=3.2, col='blue') # made up






# using mean and variance (tweedie)
Y.abun2 <- tsab.norm[,-1]
std.abund2 <- apply(Y.abun2, 2, function(vecc) sqrt(sum(vecc^2)))
log.var2 <- log(std.abund2^2)

meanvar_coefs <- summary(lm(log.var2~log.mean.abund2))$coefficients[,1] # 8.834982 1.663457 
meanvar_coefs #7.359187        1.538918 

int <- 7.359187 
slope <- 1.538918 
optimal.transformation.of.mean <- exp(-mean.ab^slope/exp(-int)) #theoretically optimal: see email

pdf("/workdir/users/fnn3/twins_uk/taxa/plots/tweedie_transformation_mean_plot.pdf", height=5, width=10)
ggplot(data.frame('x'=optimal.transformation.of.mean, 'y'=pofzeros.ab , 'col'=log.var2), aes(x=x, y=y, colour=col))+
  geom_point(size=0.5, alpha=.5) + 
  xlab("Expected proportion of zeros") + ylab("Observed proportion of zeros") +
  scale_color_continuous(name="log(variance)")
dev.off()

```




##Plotting the species abundances to check for Tweedie. These are all on unnormalized data
```{r}
#With no normalization, just raw counts
msab <- apply(sab2, 1, function(row)
  c("mean" = mean(row) , "pofzeros" = sum(row==0)/length(row))
  )   #mean abundance and proportion of zeros for each species

mean.ab <- msab[1,]
pofzeros.ab <- msab[2,]
std.abund2 <- apply(sab2, 1, function(vecc) sqrt(sum(vecc^2)))
log.var2 <- log(std.abund2^2)


plot(mean.ab, pofzeros.ab, main="Mean species abundance vs proportion of zeros") #looks ok...
plot(exp(-mean.ab^0.5/9), pofzeros.ab, main = "Tweedie expected proportion of zeros vs observed proportion of zero") #does not look right

pofnonzero <- 1 - pofzeros.ab
log.mean.abund2 <- log(mean.ab)
our.glm <- glm(pofnonzero ~ log.mean.abund2, family = binomial(link=cloglog), weights = rep(250, nrow(sab2)))
int <- our.glm$coefficients[1]
slope <- our.glm$coefficients[2]
optimal.transformation.of.mean <- exp(-mean.ab^slope/exp(-int)) #theoretically optimal: see email

pdf("/workdir/users/fnn3/twins_uk/taxa/plots/tweedie_transformation_mean_plot.pdf", height=5, width=10)
ggplot(data.frame('x'=optimal.transformation.of.mean, 'y'=pofzeros.ab , 'col'=log.var2), aes(x=x, y=y, colour=col))+
  geom_point(size=0.5, alpha=.5) + 
  xlab("Expected proportion of zeros") + ylab("Observed proportion of zeros") +
  scale_color_continuous(name="log(variance)")
dev.off()


#qplot(optimal.transformation.of.mean, pofzeros.abund2)
ggplot(data.frame('x'=optimal.transformation.of.mean, 'y'=pofzeros.ab, 'col'=log.var2), aes(x=x, y=y, colour=col))+
  geom_point(size=.5, alpha=.5)

qplot(log.mean.abund2, log.var2)+geom_abline(slope = 1, intercept = 0) + xlab("log mean") + ylab("log variance")
summary(lm(log.var2~log.mean.abund2))$coefficients[2,1]

pdf("/workdir/users/fnn3/twins_uk/taxa/plots/twins_uk_logmean_logvar.pdf", height=6, width=10)
ggplot(data.frame('x'=log.mean.abund2, 'y'=log.var2, 'col'=pofzeros.ab), aes(x=x, y=y, colour=col))+
  geom_point( alpha=.5) +
  ylab("log(var)") + xlab("log(mean)") +
  scale_color_continuous(name="Proportion of zeros")
dev.off()

plot(log.mean.abund2, log.var2, pch=20, xlim=c(0,8.5), ylim=c(0,34))#, pch=20, cex=.01)
abline(a=6, b=3.2, col='blue') # from zeros
plot(log.mean.abund2, log.var2, pch=20)#, pch=20, cex=.01)
abline(a=6, b=3.2, col='blue') # made up
```



##Plotting the species abundances to check for tweedie. normalized data, not filtered
```{r}
sabund <- read.table("/workdir/users/fnn3/twins_uk/taxa/abundance/twinsUK_species_abundance_gmeanNormalized.txt", sep="\t")
sabund2 <- sabund[,-1] #Remove character identifier, BGI_IDs
sabund2[is.na(sabund2)] <- 0  #replace NAs with 0

#### get transformations and summaries ####
m <- apply(sabund2, 2, function(col) 
   c("mean" = mean(col), "pofzeros" = sum(col==0)/length(col))
  ) #figure out mean abundance and proportion of zeroes of each gene family 

mean.sabund <- m[1,]
pofzeros.sabund <- m[2,]

# using sparsity and mean (tweedie)
spofnonzero <- 1 - pofzeros.sabund
log.mean.sabund <- log(mean.sabund)
our.glm <- glm(spofnonzero ~ log.mean.sabund, family = binomial(link=cloglog), weights = rep(250, ncol(sabund2)))
int <- our.glm$coefficients[1]
slope <- our.glm$coefficients[2]
optimal.transformation.of.mean <- exp(-mean.sabund^slope/exp(-int)) #theoretically optimal: see email

# using mean and variance (tweedie)
Y.sabund <- sabund2
std.sabund <- apply(Y.sabund, 2, function(vecc) sqrt(sum(vecc^2)))
log.svar <- log(std.sabund^2)

meanvar_scoefs <- summary(lm(log.svar~log.mean.sabund))$coefficients[,1] # int=7.36, slope= 1.54
meanvar_scoefs

# using mean and variance (negative binomial)
# note, i'd rather do this (and all others) via coordinate descent on \mu and the model parameters.. actually fitting the likelihood
#Xregr <- kronecker(diag(rep(1, ncol(abund))), rep(1, nrow(abund)))
#Yregr <- as.vector(abund)

nb_smeanvar_coefs <- c(log(1), summary(lm(log.svar~0+log.mean.sabund))$coefficients[,1])
nb_smeanvar_coefs # 0, 2.12599

#### Make plots ####

pdf("/workdir/users/fnn3/twins_uk/scca/species/plots/tweedie_transformation_mean_plot_species.pdf", height=5, width=10)
ggplot(data.frame('x'=optimal.transformation.of.mean, 'y'=pofzeros.sabund, 'col'=log.svar), aes(x=x, y=y, colour=col))+
  geom_point(size=0.5, alpha=.5) + 
  xlab("Expected proportion of zeros") + ylab("Observed proportion of zeros") +
  scale_color_continuous(name="log(variance)") +
  theme_classic()
dev.off()


#pdf("/workdir/users/fnn3/twins_uk/tweedie_mean_vs_var_plot.pdf", height=5, width=10)
ggplot(data.frame('x'=log.mean.sabund, 'y'=log.svar, 'col'=pofzeros.sabund), aes(x=x, y=y, colour=col))+
  geom_point(size=.3, alpha=.5) + 
  scale_color_continuous(name="Proportion of zeros") +
  xlab("log(mean abundance)") + ylab("log(variance)") +
  geom_abline(intercept=8, slope=1.75, col='magenta', size=1, linetype='dotted') +# made up
  theme_classic()
  #dev.off()

pdf("/workdir/users/fnn3/twins_uk/scca/species/plots/tweedie_mean_vs_var_plot_w_lines_species.pdf", height=5, width=10)
ggplot(data.frame('x'=log.mean.sabund, 'y'=log.svar, 'col'=pofzeros.sabund), aes(x=x, y=y, colour=col))+
  geom_point(size=.3, alpha=.5) + 
  scale_color_continuous(name="Proportion of zeros") +
  xlab("log(mean abundance)") + ylab("log(variance)") +
  geom_abline(intercept=0, slope=1, show.legend=T) + # poisson
  geom_abline(intercept=meanvar_scoefs[1], slope=1.80, col='red', show.legend=T) +  # from tweedie mean and var
  geom_abline(intercept=nb_smeanvar_coefs[1], slope=nb_smeanvar_coefs[2], col='blue', show.legend=T) +  # from NB1 mean and var
  theme_classic() 
dev.off()


```